CODECHECK: An open-science initiative for the independent execution of computations underlying research articles

### 2021-01-12 codecheck-paper.tex
minor formatting changes

\textbf{CODECHECK requirements are not demanding enough}:
CODECHECK by design does not require authors to provide (and sustain) an
eternally functional workflow and neither suggests a specific software stack
or practical approaches, such as research compendia
\footnote{\url{https://research-compendium.science/}},
even if they are very valuable.
Creating something that anyone can reproduce has been called a 
fool's errand
\footnote{\url{https://twitter.com/DougBlank/status/1135904909663068165?s=09}}
and we tend to agree.
However, the package of data, code, and documentation
collaboratively created by authors and codecheckers is a snapshot of a 
working analysis that greatly increases the likelihood of a successful 
reproduction and possibility to extensions by third parties, with access
to suitable resources and matching skill set, in the future.
Concrete implementations of CODECHECK workflows, especially for specific
disciplines, may reify much more helpful guidelines for authors how to
create reproducibility packages.
The author-friendly ``low bar'' should not stay low forever, but cultural
change takes time and the encouragement and guidance that CODECHECK,
as part of the widely accepted peer review concept, can provide may
eventually allow to raise the bar as high as possible, e.g., with
executable research compendia \cite{nust_opening_2017}
or continuous analysis \cite{beaulieu-jones_reproducibility_2017-1}.
However, considering that missing artefacts and lack of documentation
have repeatedly been identified as main barriers to reproducibility
(e.g., \cite{stagge_assessing_2019,nust_improving_2020}),
we would not underestimate the power of a simple check.

It should be noted that a codechecker is not taking the same roles as the
\emph{statistical reviewer}, as it is practices by some journals in the 
biomedical domain (cf.~\cite{petrovecki_role_2009,greenwood_how_2015}).
The statistical reviewer actually evaluates the appropriateness of
statistical methods \cite{greenwood_how_2015} and can support topical
reviewers if, e.g., complex methods or sophisticated variants of statistical
tests are applied \cite{petrovecki_role_2009}.
The codechecker may go equally deep into the review, if they have, by 
intention of by chance, the required expertise and the required time, 
but they are by no means expected to be able to provide
such a level of scrutiny. We can imagine a tiered workflow where the 
codechecker, just as the conventional reviewer, could recommend a detailed
statistical review to the editor as they might come across upon different
hints.
%The CODECHECK provides a certain level of safety with regard to the inherent
%problems that a computational workflow might have. We want to capture the
%obvious errors, but ... https://www.quora.com/What-is-the-difference-between-safety-and-security

@@

\textbf{Handling failure}: We have no established process for the case
that a reproduction fails. This may be the sign of a pre-selection bias
of the examples, as so far all our codechecks succeeded.
In case of a journal adopting CODECHECK for all submissions, the open 
question remains as what to do in this case will have to be answered.
At the start, we doubt publicly reporting failures (i.e., the code 
would not run) will increase overall reproducibility and recommend not
to share the outcome publicly but to share the latest draft of the 
certificate with the authors.
A negative comment in a CODECHECK certificate or a failed check does
not necessarily mean the paper or the science is bad
(cf. discussion on negative comments in \cite{everythinghertz123}).
Therefore, one must be careful as to the impact of acceptance by authors 
and the concerns of volunteers this approach, or a contrary one, has.
Private feedback for failures and public acknowledgement/encouragement
for successes seems to be the way to go.
% eLife's new model (see https://osf.io/9cftx/) will publish reports evantually, e.g., after preprint has been accepted elsewhere
\cite{Rosenthal2016b} discuss incentives for different actors around the
implementation of reproducibility. We see CODECHECK as one mean for
organisations to invest in reproducibility by creating incentives
until reproducible computations become the norm.

@@

\textbf{Not possible for my data}: 
Solutions for proprietary and sensitive data exists.
Authors can provide synthetic data (cf. \cite{shannon_opening_2018}), some
data can effectively be redacted \cite{oloughlin_data_2015}, and publishers or
independent entities can provide infrastructure for sharing data and workflows
confidentially \cite{perignon_certify_2019} or with access to derived results
but not raw data \cite{shannon_opening_2018},
i.e., data enclaves \cite{foster_research_2018},
or domains of reproducibility \cite{harris_more_2017})

@@

\textbf{Checking multiple times?} If you check a preprint before peer
review or a submission to a journal with non-public review, it might
need checking again as it is modified during peer review.  This is in
part unavoidable, and as happened to us (e.g., certificate 2020-012
\cite{cert-2020-012} has been revised). This can also be desired, if
the interaction between author, reviewer, and codechecker lead to
improvements.  Since checking the manuscript the second time is likely
to be much less work than the first time, this is largely mitigated by
open and regular communication.  The higher the reproducibility and
automation, the more aspects of a CODECHECK may eventually be
supported by automation, though we believe the human codecheckers
should remain the decision maker.  Automatically detecting if results
are ``the same'' is a question intrinsically linked with each unique
process and quite hard to realise for common outputs, such as plots or
maps.  Nevertheless, continuous integration (CI) infrastructure may
help codecheckers, and, without much further work, use scientific
workflows as detectors for when a piece of software breaks if a
workflow is set up to use the latest versions of tools.

As a principle, we do not require results to be exactly the same for a
codecheck to pass, simply that the code runs and generates the output
files that the author claims. Stochastic simulations mean that often
we cannot get exactly the same, or even the same versions of libraries
can generate outputs that differ by operating system
\cite{Gronenschild2012-pp}.

\begin{itemize}
\item modelDB principle: minimum of one figure should be reproducibe.
  Example certificate 2020-016 (I think) which had only a small subset
  of the possible figures. Still useful and worth sharing.
\end{itemize}

>>>>

\textbf{CODECHECK requirements are not demanding enough}:
CODECHECK by design does not require authors to provide (and sustain) an
eternally functional workflow and neither suggests a specific software stack
or practical approaches, such as research compendia
\footnote{\url{https://research-compendium.science/}},
even if they are very valuable.
Creating something that anyone can reproduce has been called a 
fool's errand
\footnote{\url{https://twitter.com/DougBlank/status/1135904909663068165?s=09}}
and we tend to agree.
However, the package of data, code, and documentation
collaboratively created by authors and codecheckers is a snapshot of a 
working analysis that greatly increases the likelihood of a successful 
reproduction and possibility to extensions by third parties, with access
to suitable resources and matching skill set, in the future.
Concrete implementations of CODECHECK workflows, especially for specific
disciplines, may reify much more helpful guidelines for authors how to
create reproducibility packages.
The author-friendly ``low bar'' should not stay low forever, but cultural
change takes time and the encouragement and guidance that CODECHECK,
as part of the widely accepted peer review concept, can provide may
eventually allow to raise the bar as high as possible, e.g.,~with
executable research compendia \cite{nust_opening_2017}
or continuous analysis \cite{beaulieu-jones_reproducibility_2017-1}.
However, considering that missing artefacts and lack of documentation
have repeatedly been identified as main barriers to reproducibility
(e.g.,~\cite{stagge_assessing_2019,nust_improving_2020}),
we would not underestimate the power of a simple check.

It should be noted that a codechecker is not taking the same roles as the
\emph{statistical reviewer}, as it is practices by some journals in the 
biomedical domain (cf.~\cite{petrovecki_role_2009,greenwood_how_2015}).
The statistical reviewer actually evaluates the appropriateness of
statistical methods \cite{greenwood_how_2015} and can support topical
reviewers if, e.g.,~complex methods or sophisticated variants of statistical
tests are applied \cite{petrovecki_role_2009}.
The codechecker may go equally deep into the review, if they have, by 
intention of by chance, the required expertise and the required time, 
but they are by no means expected to be able to provide
such a level of scrutiny. We can imagine a tiered workflow where the 
codechecker, just as the conventional reviewer, could recommend a detailed
statistical review to the editor as they might come across upon different
hints.
%The CODECHECK provides a certain level of safety with regard to the inherent
%problems that a computational workflow might have. We want to capture the
%obvious errors, but ... https://www.quora.com/What-is-the-difference-between-safety-and-security

@@

\textbf{Handling failure}: We have no established process for the case
that a reproduction fails. This may be the sign of a pre-selection bias
of the examples, as so far all our CODECHECKs succeeded.
In case of a journal adopting CODECHECK for all submissions, the open 
question remains as what to do in this case will have to be answered.
At the start, we doubt publicly reporting failures (i.e.,~the code 
would not run) will increase overall reproducibility and recommend not
to share the outcome publicly but to share the latest draft of the 
certificate with the authors.
A negative comment in a CODECHECK certificate or a failed check does
not necessarily mean the paper or the science is bad
(cf. discussion on negative comments in \cite{everythinghertz123}).
Therefore, one must be careful as to the impact of acceptance by authors 
and the concerns of volunteers this approach, or a contrary one, has.
Private feedback for failures and public acknowledgement/encouragement
for successes seems to be the way to go.
% eLife's new model (see https://osf.io/9cftx/) will publish reports evantually, e.g.,~after preprint has been accepted elsewhere
\cite{Rosenthal2016b} discuss incentives for different actors around the
implementation of reproducibility. We see CODECHECK as one mean for
organisations to invest in reproducibility by creating incentives
until reproducible computations become the norm.

@@

\textbf{Not possible for my data}: 
Solutions for proprietary and sensitive data exists.
Authors can provide synthetic data (cf. \cite{shannon_opening_2018}), some
data can effectively be redacted \cite{oloughlin_data_2015}, and publishers or
independent entities can provide infrastructure for sharing data and workflows
confidentially \cite{perignon_certify_2019} or with access to derived results
but not raw data \cite{shannon_opening_2018},
i.e.,~data enclaves \cite{foster_research_2018},
or domains of reproducibility \cite{harris_more_2017})

@@

\textbf{Checking multiple times?} If you check a preprint before peer
review or a submission to a journal with non-public review, it might
need checking again as it is modified during peer review.  This is in
part unavoidable, and as happened to us (e.g.,~certificate \texttt{2020-012}
\cite{cert-2020-012} has been revised). This can also be desired, if
the interaction between author, reviewer, and codechecker lead to
improvements.  Since checking the manuscript the second time is likely
to be much less work than the first time, this is largely mitigated by
open and regular communication.  The higher the reproducibility and
automation, the more aspects of a CODECHECK may eventually be
supported by automation, though we believe the human codecheckers
should remain the decision maker.  Automatically detecting if results
are ``the same'' is a question intrinsically linked with each unique
process and quite hard to realise for common outputs, such as plots or
maps.  Nevertheless, continuous integration (CI) infrastructure may
help codecheckers, and, without much further work, use scientific
workflows as detectors for when a piece of software breaks if a
workflow is set up to use the latest versions of tools.

As a principle, we do not require results to be exactly the same for a
CODECHECK to pass, simply that the code runs and generates the output
files that the author claims. Stochastic simulations mean that often
we cannot get exactly the same, or even the same versions of libraries
can generate outputs that differ by operating system
\cite{Gronenschild2012-pp}.

\begin{itemize}
\item TODO SJE: modelDB principle: minimum of one figure should be reproducibe.
  Example certificate \texttt{2020-016} (I think) which had only a small subset
  of the possible figures. Still useful and worth sharing.
\end{itemize}
